---
title: "Rideshare in Chicago During COVID-19"
author: "Nathalie P. Voorhees Center for Neighborhood and Community Improvement at UIC"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    toc_depth: 4
editor_options: 
  chunk_output_type: console
---

<style>
#exec-summ{
  height: 1375px;
  padding: 10px;
}

.exec-summ-column-left{
  float: left;
  height: 100%;
  width: 33%;
  text-align: left;
}
.exec-summ-column-right{
  float: right;
  height: 100%;
  width: 66%;
  text-align: left;
}

@media print {
  #TOC { 
    display: none; 
  }
  @page { 
    size: 8.5in 11in;
    margin: 0;
  }
}

</style>

```{r setup, include=FALSE}

# Load libraries
library(tidyverse);library(lubridate);library(janitor)
library(sf);library(stplanr);library(leaflet)
library(leaflet.extras);library(tmap);library(cartography)
library(plotly);library(knitr);library(here);library(ggspatial)
library(kableExtra);library(viridis);library(scales)
library(hrbrthemes);library(ggthemes);library(RPostgres)
library(RPostgreSQL);library(DBI);library(config)

# Set working directory and options
setwd(here::here())
options(scipen = 999, stringsAsFactors = FALSE, dplyr.summarise.inform = FALSE)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE,
                      fig.width = 10, fig.asp = 0.618,
                      fig.align = "center", out.width = "100%")

tmap_mode("view")
tmap_options(basemaps = c(Light = "CartoDB.Positron",
                          Dark = "CartoDB.DarkMatter",
                          OSM = "OpenStreetMap"))

# Plotly axis function
layout_ggplotly <- function(gg, x = -0.06, y = -0.06){
  # The 1 and 2 goes into the list that contains the options for the x and y axis labels respectively
  gg[['x']][['layout']][['annotations']][[1]][['y']] <- x
  gg[['x']][['layout']][['annotations']][[2]][['x']] <- y
  gg
}

```

```{r base data import}

# Connect to database
con <- RPostgres::dbConnect(RPostgres::Postgres(), 
                            host = config::get("host"),
                            port = config::get("port"),
                            dbname = config::get("dbname"),
                            user = config::get("user"), 
                            password = config::get("password"))

# Load in other data
# trips <- dbGetQuery(con, "SELECT * FROM public.trips LIMIT 100;")
# trips <- dbReadTable(con, "public.trips")

# analysis_trips <- dbGetQuery(con, "SELECT * FROM analysis.base_ytd LIMIT 100;")
# analysis_trips <- dbReadTable(con, "analysis.base_ytd")

# Drivers
# drivers <- dbGetQuery(con, "SELECT * FROM public.drivers LIMIT 100;")
# drivers <- dbReadTable(con, "public.drivers")

# Vehicles
# vehicles <- dbGetQuery(con, "SELECT * FROM public.vehicles LIMIT 100;")
# vehicles <- dbReadTable(con, "public.vehicles")

```

```{r daily trips import}

# Import data
daily_trips <- dbGetQuery(con, "SELECT * FROM analysis.daily_trips")

# Clean and tidy
daily_trips <- daily_trips %>%
  mutate("year" = year(date),
         "day" = yday(date),
         "date_x" = date %>% 
           as.character(.) %>%
           str_replace("^\\d{4}","2000") %>% 
           as_date(.),
         "weekday" = wday(date, label = TRUE),
         "trips" = as.numeric(trips)) 

```

```{r daily trips ggplot}

# Plot information
daily_trips_plot_title <- daily_trips %>%
  group_by(year) %>%
  summarize(trips = sum(trips)) %>%
  pivot_wider(everything(), names_from = "year", values_from = "trips") %>% 
  mutate("change" = `2020`-`2019`,
         "change_pct" = (change/`2019`),
         "title_text" = paste0(as.character(comma(-change)), 
                               " (", percent(change_pct), ")"))

# Daily trips YoY plot
gg_daily_trips <- daily_trips %>% 
  group_by(year) %>% 
  arrange(date_x) %>%
  mutate(trips_agg = cumsum(trips)) %>% 
  ggplot(aes(x = date_x, y = trips_agg, 
             color = as.character(year),
             text = paste0("Date: ", paste0(format(date_x, "%m/%d"),
                                            "/", as.character(year)),
                           "<br>Agg. Trips: ", scales::comma(trips_agg)))) +
  geom_line(group = 1, size = 1.5) +
  geom_vline(xintercept = as.numeric(as.Date("2000-03-21")), 
             linetype = 4, color = "black", size = 1) +
  scale_x_date(date_breaks = "1 month", 
               date_labels = "%b") +
  scale_y_continuous(labels = comma,
                     limits = c(0, NA)) +
  labs(title = paste0("Compared to 2019, rideshare trips fell by ",
                      daily_trips_plot_title$title_text, " in 2020"),
       x = "Day of the Year", y = "Aggregate Trips",
       color = "Year") +
  scale_color_ipsum() +
  theme_ipsum()

```

```{r daily trips kable data}

# Pre/Post COVID table
daily_trips_table <- daily_trips %>%
  mutate("covid_group" = ifelse(date_x >= "2000-03-21", 
                                "Post-COVID (3/21-12/31)", 
                                "Pre-COVID (1/1-3/21)")) %>% 
  mutate("covid_group" = factor(covid_group, 
                                levels = c("Pre-COVID (1/1-3/21)", 
                                           "Post-COVID (3/21-12/31)"))) %>% 
  group_by(covid_group, year) %>%
  summarize(trips = sum(trips)) %>%
  ungroup() %>% 
  pivot_wider(names_from = "year", 
              values_from = "trips") %>% 
  adorn_totals("row") %>% 
  mutate("change" = `2020`-`2019`,
         "change_pct" = (change/`2019`)) %>% 
  mutate(across(c(`2019`, `2020`, change), comma),
         "change_pct" = percent(change_pct)) %>% 
  select("Period" = covid_group, `2019`, `2020`, 
         "Change" = change, "Change (%)" = change_pct)

```

```{r trips around stay at home ggplot}

# Trips per day around Stay-at-Home Order
gg_trips_per_day <- daily_trips %>%
  filter(month(date, label = TRUE) %in% c("Feb", "Mar")) %>% 
  group_by(year) %>% 
  arrange(date_x) %>%
  ggplot(aes(x = date_x, y = trips, color = as.character(year),
             text = paste0("Date: ", paste0(format(date_x, "%m/%d"),
                                            "/", as.character(year)),
                           "<br>Trips: ", scales::comma(trips)))) +
  geom_line(group = 1, size = 1.5) +
  geom_vline(xintercept = as.numeric(as.Date("2000-03-21")), 
             linetype = 4, color = "black", size = 1) +
  scale_x_date(date_breaks = "1 week", 
               date_labels = "%b-%d") +
  scale_y_continuous(labels = comma,
                     limits = c(0, NA)) +
  labs(title = "Trips per Day fell before the Stay-at-Home Order went into effect,\n peaking at 365,000 trips on Saturday, March 14th, 2020",
       x = "Day of the Year", y = "Trips per Day",
       color = "Year") +
  scale_color_ipsum() +
  theme_ipsum()

```

```{r monthly trips import}

# Import data
monthly_trips <- dbGetQuery(con, "SELECT * FROM analysis.monthly_trips")

# Clean and tidy
monthly_trips <- monthly_trips %>% 
  mutate("year" = year(month),
         "date_x" = floor_date(month, unit = "month") %>% 
           as.character(.) %>%
           str_replace("^\\d{4}","2000") %>% 
           as_date(.),
         "month" = month(month, label = TRUE, abbr = TRUE),
         "trips" = as.numeric(trips)) 

```

```{r monthly trips ggplot}

# Monthly trips YoY plot
gg_monthly_trips <- monthly_trips %>% 
  ggplot(aes(x = date_x, y = trips, fill = as.character(year),
             text = paste0("Date: ", paste0(format(date_x, "%M"),
                                            "/", as.character(year)),
                           "<br>Trips: ", scales::comma(trips)))) +
  geom_bar(stat = "identity", color = "black",
           position = position_dodge(width = 25), width = 20) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  scale_y_continuous(labels = comma) +
  labs(title = "Trips per Month declined significantly in March, bottoming out at 1.5M in April.\nHowever, rideshare trips rebounded to 3.6M in the fall.",
       x = "Month", y = "Trips",
       fill = "Year") +
  scale_fill_ipsum() +
  theme_ipsum()

```

```{r day of week trips ggplot}

# Day of week YoY plot
gg_day_of_week_trips <- daily_trips %>%
  mutate("covid_group" = ifelse(date_x >= "2000-03-21", 
                                "Post-COVID", 
                                "Pre-COVID")) %>% 
  mutate("covid_group" = factor(covid_group, 
                                levels = c("Pre-COVID", 
                                           "Post-COVID"))) %>% 
  group_by(covid_group, year, weekday) %>%
  summarize("trips" = sum(trips)) %>%
  ungroup(weekday) %>%
  mutate("trips_pct" = trips / sum(trips)) %>% 
  ungroup() %>% 
  ggplot(aes(x = weekday, y = trips_pct, 
             fill = as.character(year),
             text = paste0("Day: ", paste0(weekday, "-", 
                                            as.character(year)),
                           "<br>Trips: ", paste0(scales::comma(trips)),
                           " (", scales::percent(trips_pct,
                                                 accuracy = 0.1), ")"))) +
  geom_bar(stat = "identity", color = "black") +
  scale_y_continuous(labels = percent) +
  labs(title = "Trips by Day of Week (%), Pre/Post-COVID",
       x = "Day of Week", y = "Trips (%)",
       fill = "Year") +
  facet_grid(covid_group ~ year) +
  scale_fill_ipsum() +
  theme_ipsum()

```

```{r hour of day trips import}

# Import data
hourly_trips <- dbGetQuery(con, "SELECT * FROM analysis.hourly_trips")

# Clean and tidy (Sunday is 0 in dow)
hourly_trips <- hourly_trips %>%
  mutate("day_of_week" = ifelse(day_of_week %in% c(6, 0),
                                "Weekend", "Weekday"),
         "hour_of_day" = paste0(hour_of_day, ":00:00"),
         "hour_of_day" = hour_of_day %>% hms(),
         "trips" = as.numeric(trips)) %>% 
  group_by(year, covid_group, day_of_week, hour_of_day) %>% 
  summarize("trips" = sum(trips)) %>% 
  ungroup(hour_of_day) %>%
  mutate("trips_pct" = trips / sum(trips)) %>% 
  ungroup()

```

```{r hour of day trips ggplot}

# Hour of Day YoY plot
gg_hour_of_day_trips <- hourly_trips %>%
  mutate("covid_group" = factor(covid_group, 
                                levels = c("Pre-COVID", 
                                           "Post-COVID"))) %>% 
  ggplot(aes(x = hour_of_day, y = trips_pct, 
             
             linetype = day_of_week,
             group = interaction(as.character(year), 
                                 day_of_week),
             text = paste0("Hour of Day: ", hour_of_day,
                           "<br>Day of Week: ", paste0(day_of_week, "-", 
                                            as.character(year)),
                           "<br>Trips: ", paste0(scales::comma(trips)),
                           " (", scales::percent(trips_pct,
                                                 accuracy = 0.1), ")"))) +
  aes(color = as.character(year), show.guide = FALSE) +
  geom_line(size = 1.5) +
  scale_x_time(breaks = breaks_width("6 hours"), label = label_time("%l %p")) +
  scale_y_continuous(labels = percent,
                     limits = c(0, NA)) +
  labs(title = "Trips by Hour of Day (%), Pre/Post-COVID",
       x = "Hour of Day", y = "Trips (%)",
       color = "Year", linetype = "Day of Week") +
  facet_grid(covid_group ~ year) +
  scale_fill_ipsum() +
  theme_ipsum()

```

```{r trip characteristics import}

# Import data
trip_chars_avg <- dbGetQuery(con, "SELECT * FROM analysis.tnp_trip_chars") %>% 
  mutate("covid_group" = factor(covid_group, 
                                levels = c("Pre-COVID", 
                                           "Post-COVID")),
         "trips" = as.numeric(trips)) %>% 
  arrange(covid_group, year)

```

```{r distance kable data}

# Average distance
# Pre/Post COVID table
trip_chars_distance <- trip_chars_avg %>%
  select(covid_group, year, trip_miles) %>% 
  pivot_wider(names_from = "year", 
              values_from = "trip_miles") %>% 
  mutate("change" = `2020`-`2019`,
         "change_pct" = (change/`2019`)) %>% 
  mutate(across(c(`2019`, `2020`, change), comma),
         "change_pct" = percent(change_pct)) %>% 
  select("Period" = covid_group, `2019`, `2020`, 
         "Change" = change, "Change (%)" = change_pct)

```

```{r duration kable data}

# Average time - minutes
# Pre/Post COVID table
trip_chars_duration <- trip_chars_avg %>%
  mutate("trip_minutes" = trip_seconds / 60) %>% 
  select(covid_group, year, trip_minutes) %>% 
  pivot_wider(names_from = "year", 
              values_from = "trip_minutes") %>% 
  mutate("change" = `2020`-`2019`,
         "change_pct" = (change/`2019`)) %>% 
  mutate(across(c(`2019`, `2020`, change), comma),
         "change_pct" = percent(change_pct)) %>% 
  select("Period" = covid_group, `2019`, `2020`, 
         "Change" = change, "Change (%)" = change_pct)

```

```{r fares import}

# Import data
fares <- dbGetQuery(con, "SELECT * FROM analysis.tnp_fares") %>% 
  mutate("covid_group" = factor(covid_group, 
                                levels = c("Pre-COVID", 
                                           "Post-COVID")),
         "trips" = as.numeric(trips))

# Distribution of fares
fares_pct <- fares %>%
  group_by(covid_group, year) %>% 
  arrange(year, covid_group, fare) %>% 
  mutate("trips_pct" = trips / sum(trips),
         "trips_pct_agg" = cumsum(trips_pct)) %>% 
  ungroup()

# 95% of trips in all periods <$30 dollars
fares_tail <- fares_pct %>% 
  group_by(covid_group, year) %>% 
  filter(trips_pct_agg < 0.95) %>% 
  arrange(desc(fare)) %>% 
  slice(1) %>% 
  ungroup()

```

```{r fares kable data}

# Average fares
# Pre/Post COVID table
trip_chars_fares <- trip_chars_avg %>%
  select(covid_group, year, fare) %>% 
  pivot_wider(names_from = "year", 
              values_from = "fare") %>% 
  mutate("change" = `2020`-`2019`,
         "change_pct" = (change/`2019`)) %>% 
  mutate(across(c(`2019`, `2020`, change), dollar),
         "change_pct" = percent(change_pct)) %>% 
  select("Period" = covid_group, `2019`, `2020`, 
         "Change" = change, "Change (%)" = change_pct)

```

```{r fares ggplot, warning=FALSE}

# Histogram
gg_fares <- fares %>%
  ggplot(aes(x = fare, y = trips, fill = as.character(year),
             text = paste0("Fare: ",scales::dollar(fare),
                           "<br>Trips: ", scales::comma(trips)))) +
  geom_bar(stat = "identity", color = "black") +
  geom_vline(data = trip_chars_avg, 
             mapping = aes(xintercept = fare),
             linetype = "dashed", size = 1) +
  # Set limit to ~$30 to eliminate long tail and focus on bulk of trips
  scale_x_continuous(labels = dollar_format(prefix = "$"),
                     limits = c(0, max(fares_tail$fare))) +
  scale_y_continuous(labels = comma) +
  labs(title = "Fare Distribution to the Nearest $2.50, Pre/Post-COVID",
       subtitle = "95% of trip fares were less than $30 in each period",
       x = "Fare", y = "Trips",
       fill = "Year") +
  facet_grid(covid_group ~ year) +
  scale_fill_ipsum() +
  theme_ipsum()

```

```{r od spatial import}

# Import data
origin_tracts_comp <- dbGetQuery(con, "SELECT * FROM analysis.pickup_tracts_comp")
destination_tracts_comp <- dbGetQuery(con, "SELECT * FROM analysis.dropoff_tracts_comp")
sf_tracts <- st_read(con, layer = "spatial_tracts")
sf_ccas <- st_read(con, layer = "spatial_community_areas")

```

```{r essential workers definition}

# Create essential worker definition
# National averages:
#     -Essential: 23.8%
#     -Healthcare: 9.5%
# City of Chicago:
#     -Essential: 334,381 (24.5%)
#     -Essential Non Police/Fire: 293,575 (21.5%)
#     -Healthcare: 118,787 (8.7%)
# Tract median (50th):
# Share of city employment in occupation
#     -Essential: 353 (0.11%)
#     -Essential Non Police/Fire: 314 (0.11%)
#     -Healthcare: 115 (0.10%)
# Tract top 20 percent, first quintile (80th):
# Share of city employment in occupation
#     -Essential: 610 (0.18%), combined 136,540 (40.8%)
#     -Essential Non Police/Fire: 532 (0.18%), combined 121,060 (41.5%)
#     -Healthcare: 214 (0.18%), combined 55,398 (46.6%)

# Import data
acs_data_profiles <- dbGetQuery(con, "SELECT * FROM public.acs_data_profiles")

# Chicago 
acs_data_chi <- acs_data_profiles %>% 
  filter(table_name == "S2401") %>% 
  select(GEOID, variable, estimate) %>% 
  pivot_wider(names_from = "variable",
              values_from = "estimate") %>%
  select(-GEOID) %>% 
  summarize(across(where(is.numeric), ~ sum(.x, na.rm = TRUE))) %>% 
  mutate("essential" = select(., S2401_C01_015:S2401_C01_025) %>% 
           rowSums(na.rm = TRUE),
         "essential_pct" = essential / S2401_C01_001,
         "essential_non_police_fire" = select(., S2401_C01_015:S2401_C01_025,
                                              -S2401_C01_020) %>% 
           rowSums(na.rm = TRUE),
         "essential_non_police_fire_pct" = essential_non_police_fire / S2401_C01_001,
         "health_all" = select(., S2401_C01_015:S2401_C01_019) %>% 
           rowSums(na.rm = TRUE),
         "health_all_pct" = health_all / S2401_C01_001) %>% 
  rename("total" = S2401_C01_001,
         "health_pract" = S2401_C01_015,
         "health_support" = S2401_C01_019,
         "police_fire" = S2401_C01_020,
         "food_service" = S2401_C01_023,
         "custodial" = S2401_C01_024,
         "personal_service" = S2401_C01_025)

# Tract level, calculate share of city employment by occupation
acs_data_tracts <- acs_data_profiles %>% 
  filter(table_name == "S2401") %>% 
  select(GEOID, variable, estimate) %>% 
  pivot_wider(names_from = "variable",
              values_from = "estimate") %>% 
  mutate("essential" = select(., S2401_C01_015:S2401_C01_025) %>% 
           rowSums(na.rm = TRUE),
         "essential_share_chi_pct" = essential / acs_data_chi$essential,
         "essential_non_police_fire" = select(., S2401_C01_015:S2401_C01_025,
                                              -S2401_C01_020) %>% 
           rowSums(na.rm = TRUE),
         "essential_non_police_fire_pct" = essential_non_police_fire / acs_data_chi$essential_non_police_fire,
         "health_all" = select(., S2401_C01_015:S2401_C01_019) %>% 
           rowSums(na.rm = TRUE),
         "health_all_share_chi_pct" = health_all / acs_data_chi$health_all) %>%
  
  rename("total" = S2401_C01_001,
         "health_pract" = S2401_C01_015,
         "health_support" = S2401_C01_019,
         "police_fire" = S2401_C01_020,
         "food_service" = S2401_C01_023,
         "custodial" = S2401_C01_024,
         "personal_service" = S2401_C01_025)

# Essential workers by tract
essential_workers_sf <- left_join(sf_tracts %>% select(GEOID),
                                  acs_data_tracts,
                                  by = "GEOID")

```

```{r od tracts data}

# Join trip data to spatial, reconfigure
# Origin-Destination
od_tracts <- 
  # Origin and Destination
  full_join(origin_tracts_comp %>% 
              rename("origin_trips" = trips),
            destination_tracts_comp %>%
              rename("destination_trips" = trips),
            by = c("pickup_census_tract" = "dropoff_census_tract",
                   "year" = "year",
                   "covid_group" = "covid_group")) %>% 
  mutate("origin_trips" = as.numeric(origin_trips),
         "destination_trips" = as.numeric(destination_trips),
         "covid_group" = factor(covid_group, 
                                levels = c("Pre-COVID", 
                                           "Post-COVID"))) %>% 
  select("GEOID" = pickup_census_tract, year, covid_group,
         origin_trips, destination_trips) %>% 
  group_by(year, covid_group) %>% 
  mutate("origin_rank" = percent_rank(origin_trips),
         "destination_rank" = percent_rank(destination_trips)) %>% 
  ungroup(year, covid_group) %>%
  # Restrict to Post-COVID
  filter(covid_group == "Post-COVID") %>% 
  # Join to spatial data, remove tracts outside Chicago
  left_join(sf_tracts %>% select(GEOID),
            .,
            by = "GEOID") %>% 
  # Reconfigure and format
  st_as_sf() %>% 
  arrange(GEOID, covid_group, year)


# Join trip data to spatial, reconfigure
# Origin-Destination
od_tracts_rank_change <- od_tracts %>%
  st_drop_geometry() %>% 
  pivot_wider(names_from = year,
              values_from = c(origin_trips, origin_rank,
                              destination_trips, destination_rank)) %>%
  mutate("origin_rank_change" = round((origin_rank_2020 -
                                         origin_rank_2019) * 100, 0),
         "destination_rank_change" = round((destination_rank_2020 -
                                              destination_rank_2019) * 100, 0)) %>% 
  # Join to spatial data, remove tracts outside Chicago
  left_join(sf_tracts %>% select(GEOID),
            .,
            by = "GEOID") %>% 
  # Reconfigure and format
  st_as_sf() 

```

```{r origin tracts tmap data, warning=FALSE}

# Origin quantile map
tmap_origin_facet <-
  tm_shape(od_tracts %>%
             mutate("GEOID" = paste0("GEOID: ", GEOID)),
           name = "Trips by Origin") +
  tm_fill(col = "origin_trips", title = "Trips by Origin",
          alpha = 0.8,
          style = "log10_pretty",
          palette = "plasma",
          popup.vars = c("Trips: " = "origin_trips")) +
  tm_borders(lwd = 0.5) +
  tm_facets(by = c("covid_group", "year"),
            nrow = 1, ncol = 2,
            free.scales = FALSE,
            drop.units = TRUE) +
  # Essential workers, non-police/fire, top 20%
  tm_shape(essential_workers_sf %>% 
             filter(essential_non_police_fire_pct >= 
                      quantile(essential_non_police_fire_pct,
                               0.80,
                               na.rm = TRUE)),
           name = "Share of Chicago Essential Workers<br>Top 20 Percent of Tracts") +
  tm_borders(col = "black", lwd = 0.75) +
  # Layout
  tm_layout(scale = 2.75) +
  tm_view(view.legend.position = c("left", "bottom"))

# Origin change in percent rank map
tmap_origin_rank_change <-
  tm_basemap(leaflet::providers$CartoDB.DarkMatter) +
  tm_shape(od_tracts_rank_change %>% 
             mutate("GEOID" = paste0("GEOID: ", GEOID)),
           name = "Change in Percentile<br>by Trip Origin") +
  tm_fill(col = "origin_rank_change",
          title = "Change in Percentile<br>by Trip Origin",
          alpha = 0.9, breaks = c(-Inf, -25, -15, -5,
                                  5, 15, 25, Inf),
          palette = "RdYlGn", midpoint = 0,
          popup.vars = c("Change in Percentile: " = "origin_rank_change")) +
  tm_borders(lwd = 0.5) +
  # Essential workers, non-police/fire, top 20%
  tm_shape(essential_workers_sf %>% 
             filter(essential_non_police_fire_pct >= 
                      quantile(essential_non_police_fire_pct,
                               0.80,
                               na.rm = TRUE)),
           name = "Share of Chicago Essential Workers<br>Top 20 Percent of Tracts") +
  tm_borders(col = "black", lwd = 1.25) +
  # Layout
  tm_layout(title = "Change in Percentile<br>by Trip Origin") +
  tm_view(view.legend.position = c("left", "bottom"))

```

```{r destination tracts tmap data, warning=FALSE}

# Destination
tmap_destination_facet <-  
  tm_shape(od_tracts %>%
             mutate("GEOID" = paste0("GEOID: ", GEOID)),
           name = "Trips by Destination") +
  tm_fill(col = "destination_trips", 
          title = "Trips by Destination",
          alpha = 0.8, style = "log10_pretty",
          palette = "plasma",
          popup.vars = c("Trips: " = "destination_trips")) +
  tm_borders(lwd = 0.5) +
  tm_facets(by = c("covid_group", "year"),
            nrow = 1, ncol = 2,
            free.scales = FALSE,
            drop.units = TRUE) +
  # Essential workers, non-police/fire, top 20%
  tm_shape(essential_workers_sf %>% 
             filter(essential_non_police_fire_pct >= 
                      quantile(essential_non_police_fire_pct,
                               0.80,
                               na.rm = TRUE)),
           name = "Share of Chicago Essential Workers<br>Top 20 Percent of Tracts") +
  tm_borders(col = "black", lwd = 0.75) +
  # Layout
  tm_layout(scale = 2.75) +
  tm_view(view.legend.position = c("left", "bottom"))

# Destination change in percent rank map
tmap_destination_rank_change <-
  tm_basemap(leaflet::providers$CartoDB.DarkMatter) +
  tm_shape(od_tracts_rank_change %>% 
             mutate("GEOID" = paste0("GEOID: ", GEOID)),
           name = "Change in Percentile<br>by Trip Destination") +
  tm_fill(col = "destination_rank_change",
          title = "Change in Percentile<br>by Trip Destination",
          alpha = 0.9, breaks = c(-Inf, -25, -15, -5,
                                  5, 15, 25, Inf),
          palette = "RdYlGn", midpoint = 0,
          popup.vars = c("Change in Percentile: " = "destination_rank_change")) +
  tm_borders(lwd = 0.5) +
  # Essential workers, non-police/fire, top 20%
  tm_shape(essential_workers_sf %>% 
             filter(essential_non_police_fire_pct >= 
                      quantile(essential_non_police_fire_pct,
                               0.80,
                               na.rm = TRUE)),
           name = "Share of Chicago Essential Workers<br>Top 20 Percent of Tracts") +
  tm_borders(col = "black", lwd = 1.25) +
  # Layout
  tm_layout(title = "Change in Percentile<br>by Trip Destination") +
  tm_view(view.legend.position = c("left", "bottom"))

```

```{r od ccas data}

# Get trip totals by community area
# Yera, COVID group
od_ccas <- od_tracts %>% 
  st_drop_geometry() %>% 
  left_join(sf_tracts %>% 
              st_drop_geometry() %>% 
              select(GEOID, comm_area_n),
            by = "GEOID") %>% 
  left_join(sf_ccas %>% 
              select(comm_area_n, community),
            by = "comm_area_n") %>% 
  group_by(comm_area_n, community, 
           year, covid_group) %>% 
  summarize(across(ends_with("_trips"), sum)) %>% 
  mutate("trips" = origin_trips + destination_trips) %>% 
  select(comm_area_n:covid_group, trips) %>% 
  group_by(year, covid_group) %>% 
  mutate("trips_pct" = trips/sum(trips)) %>% 
  ungroup()

# Top 3 community areas
od_ccas_top3 <- od_ccas %>%
  group_by(year, covid_group) %>% 
  arrange(desc(trips_pct)) %>% 
  slice(1:3) %>% 
  summarize(across(starts_with("trips"), sum))
  

```

```{r od flows import, warning=FALSE, message=FALSE}

# Import data
# Tracts
od_pairs_tracts_raw <- dbGetQuery(con,
                                  "SELECT * FROM analysis.trip_patterns_tracts_comp") %>% 
  mutate("covid_group" = factor(covid_group, 
                                levels = c("Pre-COVID", "Post-COVID")),
         "trips" = as.numeric(trips))
# Community Areas
od_pairs_ccas_raw <- dbGetQuery(con, 
                                "SELECT * FROM analysis.trip_patterns_cca_comp") %>% 
  mutate("covid_group" = factor(covid_group, 
                                levels = c("Pre-COVID", "Post-COVID")),
         "trips" = as.numeric(trips))

# OD Pairs Intra vs Inter
# Tracts
od_pairs_tracts_intra_inter <- od_pairs_tracts_raw %>% 
  mutate("intra_inter" = case_when(pickup_census_tract == 
                                     dropoff_census_tract ~
                                     "Intra (Within Same Tract)",
                                   pickup_census_tract != 
                                     dropoff_census_tract ~
                                     "Inter (Different Tract)")) %>% 
  group_by(covid_group, year, intra_inter) %>% 
  summarize("trips" = sum(trips)) %>% 
  ungroup()

# Community Areas
od_pairs_ccas_intra_inter <- od_pairs_ccas_raw %>% 
  mutate("intra_inter" = case_when(pickup_community_area == 
                                     dropoff_community_area ~
                                     "Intra (Within Same Community Area)",
                                   pickup_community_area != 
                                     dropoff_community_area ~
                                     "Inter (Different Community Area)")) %>% 
  group_by(covid_group, year, intra_inter) %>% 
  summarize("trips" = sum(trips)) %>% 
  ungroup()

```

```{r od intra vs inter kable data}

# OD Pairs Intra vs Inter
# Pre/Post-COVID table
# Tracts
od_pairs_tracts_intra_inter_table <- od_pairs_tracts_intra_inter %>% 
  pivot_wider(names_from = c("year", "covid_group"),
              values_from = "trips") %>% 
  adorn_totals(c("row", "col")) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting()

# Community Areas
od_pairs_ccas_intra_inter_table <- od_pairs_ccas_intra_inter %>% 
  pivot_wider(names_from = c("year", "covid_group"),
              values_from = "trips") %>% 
  adorn_totals(c("row", "col")) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting()

```

```{r od airport kable data}

# Calculate airport trips
od_pairs_airports <- od_pairs_tracts_raw %>% 
  filter(pickup_census_tract != dropoff_census_tract,
         pickup_census_tract %in% sf_tracts$GEOID,
         dropoff_census_tract %in% sf_tracts$GEOID) %>% 
  mutate("airports" = ifelse(
    pickup_census_tract %in% c("17031980000", "17031980100") | dropoff_census_tract %in% c("17031980000", "17031980100"), 
    "To or From O'Hare/Midway",
    "Other Trip")) %>% 
  group_by(covid_group, year, airports) %>% 
  summarize("trips" = sum(trips)) %>% 
  ungroup()

# Airport table data
od_pairs_airports_table <- od_pairs_airports %>% 
  pivot_wider(names_from = c("year", "covid_group"),
              values_from = "trips") %>% 
  adorn_totals(c("row", "col")) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting()

```

```{r od flows calculation, warning=FALSE, message=FALSE}

# Remove airport trips, make sure in Chicago
# Convert into desire lines for mapping
# Post-COVID, 2019 vs 2020
od_pairs_ccas_comp <- od_pairs_ccas_raw %>% 
  filter(covid_group == "Post-COVID") %>% 
  filter(pickup_community_area != dropoff_community_area,
         pickup_community_area %in% sf_ccas$comm_area_n,
         dropoff_community_area %in% sf_ccas$comm_area_n) %>%
  select(pickup_community_area, dropoff_community_area,
         everything()) %>% 
  od2line(flow = ., zones = sf_ccas)

# Create a linewidth variable to scale desire lines
od_pairs_ccas_comp <- od_pairs_ccas_comp %>% 
  mutate("lwd" = scales::rescale(trips, to = c(0, 3))) %>% 
  relocate(lwd, .after = trips)

```

```{r od flows ccas tmap data}

# OD flows map (all community areas)
tmap_od_flows_ccas <- 
  tm_basemap(leaflet::providers$CartoDB.Positron) +
  tm_shape(sf_ccas, name = "Community Areas") +
  tm_borders(lwd = 0.5) +
  tm_shape(od_pairs_ccas_comp,
           name = "OD Flows") +
  tm_lines(col = "trips", title = "Trips", 
           style = "log10_pretty", palette = "plasma",
           lwd = "lwd", title.lwd = "Trips",
           scale = 15, alpha = 0.5,
           legend.lwd.show = FALSE) +
  tm_facets(by = c("covid_group", "year"),
            nrow = 1, ncol = 2,
            free.scales = TRUE,
            drop.units = TRUE) +
  tm_shape(essential_workers_sf %>% 
             filter(essential_non_police_fire_pct >= 
                      quantile(essential_non_police_fire_pct,
                               0.80,
                               na.rm = TRUE)),
           name = "Share of Chicago Essential Workers<br>Top 20 Percent of Tracts") +
  tm_borders(col = "black", lwd = 0.75) +
  tm_layout(main.title = "Rideshare Trips by Origin-Destination (OD)<br>Community Areas<br>Excluding Airports", 
            main.title.position = c("center"),
            legend.outside = FALSE,
            legend.width = 1.5,
            scale = 2.75) +
  tm_view(view.legend.position = c("left", "bottom"))

```

```{r od flows executive summary, warning=FALSE, message=FALSE}

# Executive summary OD flows
static_basemap <- getTiles(sf_ccas,
                           type = "CartoDB.PositronNoLabels",
                           crop = TRUE)

ggplot_od_flows_exec_summ <- 
  ggplot() +
  ggspatial::layer_spatial(static_basemap) +
  geom_sf(data = sf_ccas, 
          fill = NA, color = "black", size = 0.5) +
  geom_sf(data = od_pairs_ccas_comp %>% 
            filter(year == 2020,
                   covid_group == "Post-COVID") %>% 
            mutate("lwd" = (trips / 
                              mean(od_pairs_ccas_comp$trips)) * 5,
                   "alpha" = rescale(trips, 0.001, 1)) %>% 
            arrange(trips),
          aes(color = trips, lwd = lwd, alpha = alpha)) +
  scale_size(range = c(0.001, 5)) +
  scale_alpha(range = c(0, 1)) +
  scale_color_viridis(trans = "log10", option = "plasma") +
  labs(color = "Rideshare Trips\nExcluding Airports\nAfter 3/21/20\nOrigin or Destination") +
  guides(size = "none", alpha = "none") +
  theme_void() +
  theme(legend.position = c(0.2, 0.3))

```

```{r drivers import}

# Import data
drivers <- dbGetQuery(con, "SELECT * FROM analysis.drivers") %>%
  filter(year %in% c(2019, 2020)) %>%
  rename("trips" = number_of_trips) %>% 
  mutate("covid_group" = factor(covid_group, 
                                levels = c("Pre-COVID", 
                                           "Post-COVID")),
         "trips" = as.numeric(trips),
         "driver_tenure" = factor(driver_tenure, 
                                  levels = c("0-3 Months", 
                                             "3-12 Months",
                                             "Over a Year")))

# Clean and tidy
monthly_drivers <- drivers %>%
  mutate("date_x" = floor_date(month_reported, unit = "month") %>% 
           as.character(.) %>%
           str_replace("^\\d{4}","2000") %>% 
           as_date(.),
         "month_name" = month(month_reported, 
                              label = TRUE, abbr = TRUE)) 

```

```{r monthly drivers active ggplot}

# Monthly active drivers YoY plot
gg_monthly_drivers_active <- monthly_drivers %>%
  filter(trips > 0) %>% 
  group_by(date_x, year) %>% 
  summarize("drivers" = n()) %>% 
  ungroup() %>% 
  ggplot(aes(x = date_x, y = drivers, fill = as.character(year),
             text = paste0("Date: ", 
                           paste0(format(date_x, "%M"),
                                  "/", as.character(year)),
                           "<br>Active Drivers: ", 
                           scales::comma(drivers)))) +
  geom_bar(stat = "identity", color = "black",
           position = position_dodge(width = 25), width = 20) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  scale_y_continuous(labels = comma) +
  labs(title = "Active drivers per month declined starting in March 2020, bottoming out\nat 17,000 in April. The number of active drivers remained below 50% of\n2019 levels.",
       x = "Month", y = "Active Drivers (1+ Trips)",
       fill = "Year") +
  scale_fill_ipsum() +
  theme_ipsum()

```

```{r drivers status ggplot}

# Monthly active drivers YoY plot
gg_drivers_status <- monthly_drivers %>%
  mutate("driver_status" = ifelse(trips > 0,
                                   "Active (1+ Trips)",
                                   "Inactive (0 Trips)")) %>% 
  filter(!is.na(driver_status)) %>% 
  group_by(covid_group, year, driver_status) %>% 
  summarize("drivers" = n()) %>% 
  ungroup(driver_status) %>%
  mutate("drivers_pct" = drivers / sum(drivers)) %>% 
  ungroup() %>% 
  ggplot(aes(x = driver_status,
             y = drivers_pct, 
             fill = driver_status,
             text = paste0("Driver's Status: ", 
                           driver_status,
                           "<br>Trips: ",
                           paste0(scales::comma(drivers)),
                           " (", scales::percent(drivers_pct,
                                                 accuracy = 0.1), ")"))) +
  geom_bar(stat = "identity", position = "stack",
           color = "black") +
  scale_y_continuous(labels = percent) +
  labs(title = "After Stay-at-Home Order went into effect, a larger share of drivers were\ninactive (64%) compared to the same Post-COVID period in 2019 (34%).",
       x = "", y = "Drivers by Driver's Status (%)",
       fill = "Driver's Status") +
  facet_grid(covid_group ~ year) +
  scale_fill_manual(values = c(ipsum_pal()(7)[4],
                                 ipsum_pal()(7)[7])) +
  theme_ipsum()
  
```

```{r drivers tenure ggplot}

# Trips by driver tenure YoY plot
gg_drivers_tenure_trips <- monthly_drivers %>%
  filter(trips > 0, !is.na(driver_tenure)) %>% 
  group_by(covid_group, year, driver_tenure) %>%
  summarize("drivers" = n(),
            "trips" = sum(trips)) %>%
  ungroup(driver_tenure) %>%
  mutate("drivers_pct" = drivers / sum(drivers),
         "trips_pct" = trips / sum(trips)) %>% 
  ungroup() %>% 
  ggplot(aes(x = driver_tenure,
             y = trips_pct, 
             fill = driver_tenure,
             text = paste0("Driver's Tenure: ", 
                           driver_tenure,
                           "<br>Trips: ",
                           paste0(scales::comma(trips)),
                           " (", scales::percent(trips_pct,
                                                 accuracy = 0.1), ")"))) +
  geom_bar(stat = "identity", position = "stack",
           color = "black") +
  scale_y_continuous(labels = percent) +
  labs(title = "A higher percentage of trips (+7%) were made by drivers with over a year's experience\nin 2020 compared to 2019, both before and after the Stay-at-Home Order went into effect.",
       x = "", y = "Trips by Active Drivers (%)",
       fill = "Driver's Tenure") +
  facet_grid(covid_group ~ year) +
  scale_fill_ipsum() +
  theme_ipsum()
  
```

```{r drivers license ggplot}

# Trips by driver state YoY plot
gg_drivers_license_trips <- monthly_drivers %>%
  mutate("driver_license" = case_when(city == "Chicago" ~
                                         "Chicago",
                                       state == "IL" ~
                                         "Illinois",
                                       state %in% c("IN", "WI") ~
                                         "Neighboring\nStates (IN, WI)",
                                       TRUE ~
                                         "Other")) %>% 
  mutate("driver_license" = factor(driver_license,
                                    levels = c("Chicago",
                                               "Illinois",
                                               "Neighboring\nStates (IN, WI)",
                                               "Other"))) %>% 
  filter(trips > 0, !is.na(driver_license)) %>% 
  group_by(covid_group, year, driver_license) %>%
  summarize("drivers" = n(),
            "trips" = sum(trips)) %>%
  ungroup(driver_license) %>%
  mutate("drivers_pct" = drivers / sum(drivers),
         "trips_pct" = trips / sum(trips)) %>% 
  ungroup() %>% 
  ggplot(aes(x = driver_license,
             y = trips_pct, 
             fill = driver_license,
             text = paste0("Driver's License: ", 
                           driver_license,
                           "<br>Trips: ",
                           paste0(scales::comma(trips)),
                           " (", scales::percent(trips_pct,
                                                 accuracy = 0.1), ")"))) +
  geom_bar(stat = "identity", position = "stack",
           color = "black") +
  scale_y_continuous(labels = percent) +
  labs(title = "A higher percentage of trips were made by drivers with over a year's experience\nin 2020 compared to 2019, before and after the Stay-at-Home Order went into effect.",
       x = "", y = "Trips by Active Drivers (%)",
       fill = "Driver's License") +
  facet_grid(covid_group ~ year) +
  scale_fill_ipsum() +
  theme_ipsum()
  
```

<div id="exec-summ">

## Executive Summary

<div class="exec-summ-column-left">

**Rideshare trips decreased by 62 million (55%) in 2020 compared to 2019.** From late March 2020 onwards, only 27 million rideshare trips were taken in Chicago, about 60 million fewer trips compared to the same period in 2019.

**Weekday trips shifted to the afternoon, and average trip duration fell by 3 minutes (-15%), but average trip distance and fare remained unchanged.** 

**Trip origins and destinations were mostly unchanged. Excluding airport trips (6%-9% of all trips), the most common rides were short trips to and from the Loop and nearby areas.** In both 2019 and 2020, the top 3 Community Areas (The Loop, Near North Side, and Near West Side) accounted for over 40% of all rideshare trips within the city.

**While few rideshare trips were made on the South and West Sides compared to other parts of the city, rideshare trip behavior was less affected during the pandemic.** For example, one Census Tract in the West Side neighborhood of Austin was middle of the pack for trip pickups in 2019 (56th percentile). Pickups were almost cut in half in 2020, but the tract qualified as one of the more active pickup spots in the city (75th percentile).

**The number of active drivers completing 1+ trip per month decreased by 45,000 (-64%) after the Stay-at-Home Order went into effect.** This trend was also paired by a large increase in the number of inactive drivers, defined as those who completed zero trips despite being eligible to drive in Chicago that month. During this period, a higher percentage of drivers were inactive (64%) compared to 2019 (34%). A higher percentage of trips (+7%) were made by drivers with over a year's experience, both before and after the Stay-at-Home Order went into effect.

</div>

<div class="exec-summ-column-right">

```{r executive summary plot 1, fig.width=6.5, out.width = "100%"}

# Daily trips
gg_daily_trips + 
  theme_ipsum(plot_title_size = 12) + 
  theme(plot.margin = margin(0, 0, 0, 0, "cm"))
 
```


```{r executive summary plot 2, fig.width=6, fig.asp=1, out.width = "100%"}

# OD Pairs, Post-COVID 2020
ggplot_od_flows_exec_summ +
  theme(plot.margin = margin(0, 0, 0, 0, "cm"))

```

```{r executive summary plot 3, fig.width=6.5, out.width = "100%"}

# Active Drivers
gg_monthly_drivers_active + 
  theme_ipsum(plot_title_size = 12) +
  theme(plot.margin = margin(0, 0, 0, 0, "cm"))

```

</div>

</div>

\newpage

<br />
<br />

## Rideshare Data

This analysis relies on Transportation Network Provider (TNP) data for trips, drivers, and vehicles. These datasets are published on the City of Chicago's Data Portal and updated quarterly. Please see the [GitHub repository](https://github.com/sean-connelly/Chicago_TNP) for source links and download instructions.

## Trips

### Daily Trips

```{r daily trips interactive plot}

# Interactive plot
ggplotly(gg_daily_trips, tooltip = "text") %>% 
  layout(title = list(text =  paste0(
    "Compared to 2019, rideshare trips decreased by ",
    daily_trips_plot_title$title_text, " in 2020.")))

```

**This analysis uses the day that Governor Pritzker's Stay-at-Home Order went into effect (March 21st), as a way to delineate between two time frames:**

- **"Pre-COVID" (1/1-3/21) and **
- **"Post-COVID" (3/21-12/31)**

**As the data below suggest, the use of a single date to mark the onset of the pandemic is an oversimplification, but this allows for a clean comparison of rideshare trends year-over-year.** 

Prior to the Stay-at-Home Order going into effect, trips were already down by about 7 percent year-over-year. However, the bulk of the decline in rideshare occurred in the following 9 months. From late March 2020 onwards, only 27 million rideshare trips were taken in Chicago, about 60 million fewer trips compared to the same period in 2019.

```{r daily trips kable}

# Pre/Post COVID table
daily_trips_table %>%
  kbl() %>% 
  row_spec(3, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", 
                                      "condensed", "responsive"))

```

### Trips Around 3/21/20 Stay-at-Home Order

It is clear that behavior changed in March even before the Stay-at-Home Order, as trips per day decreased significantly after Saturday, March 14th, 2020, a full week before the Order went into effect. Note that the earlier fluctuations in January and February correspond to weekday versus weekend trips.

```{r trips around stay at home interactive plot}

# Interactive plot
ggplotly(gg_trips_per_day, tooltip = "text") %>% 
  layout(title = list(text = "Trips per Day fell before the Stay-at-Home Order went into effect,\npeaking at 365,000 trips on Saturday, March 14th, 2020."))

```

### Monthly Trips

```{r monthly trips interactive plot}

# Interactive plot
ggplotly(gg_monthly_trips, tooltip = "text") %>% 
  layout(title = list(text = "Trips per Month declined significantly in March, bottoming out at 1.5M in April.\nHowever, rideshare trips rebounded to 3.6M in the fall."))

```

### Trips by Day of Week and Hour of Day

#### Day of Week

While the number of trips were down significantly in the 2020 Post-COVID period compared to 2019, rideshare trips by day exhibited the same pattern across years and times periods; relatively low Monday through Wednesday before rising into the weekend and peaking Saturday.

```{r day of week trips interactive plot}

# Interactive plot
ggplotly(gg_day_of_week_trips, tooltip = "text") %>% 
  layout(title = list(text = "Trips by Day of Week (%) remained consistent, Pre/Post-COVID")) %>% 
  layout_ggplotly

```

#### Hour of Day

Breaking out trips by hour of the day, it is clear that trips were shifted outside of peak hours after the Stay-at-Home order went into effect. Weekend trips by hour (dashed line) still exhibited a late night surge, but the twin peaks of the morning and evening rush during weekdays (solid line) are less noticeable. It is clear that a higher proportion of weekday trips took place during the late afternoon hours in the 2020 Post-COVID period compared to 2019. 

```{r hour of day trips interactive plot}

# Interactive plot
#gg_hour_of_day_trips
ggplotly(gg_hour_of_day_trips, tooltip = "text") %>%
  layout(title = list(text = "Trips by Hour of Day (%) flattened on weekdays (solid line) after the\nStay-at-Home Order went into effect, suggesting some trips were\nshifted to the afternoon.",
                      y = 0.97)) %>% 
  layout_ggplotly

```

## Trip Characteristics

### Distance

Prior to the Stay-at-Home Order going into effect, trips were 6.4 miles long on average, a 9 percent increase year-over-year. From late March 2020 onwards, however, the average trip distance increased to only slightly to 6.5 miles, a 1.7 percent increase compared to the same Post-COVID period in 2019.

```{r distance kable}

# Average distance - miles
# Pre/Post COVID table
trip_chars_distance %>% 
  kbl(caption = "Average Trip Distance (Mi.)") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", 
                                      "condensed", "responsive"))

```

### Duration

While average trip distance increased slightly, the average trip duration fell significantly in 2020. After March 21, 2020, the average trip time was 15.9 minutes, a 15 percent decrease compared to the same period in 2019. This finding aligns with Manzo and Bruno's [analysis based on a sample of trips from September 2019 and September 2020](https://illinoisepi.files.wordpress.com/2021/01/ilepi-pmcr-on-demand-workers-sub-minimum-wages-final.pdf) which found a 16 percent decline in average trip duration due to a decrease in congestion.

```{r duration kable}

# Average duration - minutes
# Pre/Post COVID table
trip_chars_duration %>% 
  kbl(caption = "Average Trip Duration (Min.)") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", 
                                      "condensed", "responsive"))

```

### Fares

The average trip fare in 2020 was about $0.80 higher in the first three months of the year compared to the same period in 2019, a 7.4 percent increase. After the Stay-at-Home Order went into effect, however, the average trip fare in 2020 was about the same as in 2019.

```{r fares kable}

# Average fares
# Pre/Post COVID table
trip_chars_fares %>% 
  kbl(caption = "Average Trip Fare ($)") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", 
                                      "condensed", "responsive"))

```

```{r fares interactive plot, warning=FALSE}

# Interactive plot
ggplotly(gg_fares, tooltip = "text") %>% 
  layout(title = list(text = paste0("Fare Distribution to the Nearest $2.50, Pre/Post-COVID", 
                                    "<br>",
                                    "<sup>",
                                    "95% of trip fares were less than $30 in each period","</sup>"))) %>% 
  layout_ggplotly

```

## Spatial Distribution of Trips, Post-COVID

The spatial distribution of rideshare trips that started and ended within the City of Chicago during the Pre-COVID period (1/1-3/21) of 2019 and 2020 were very similar. **The maps below display how spatial patterns changed after the Stay-at-Home Order went into effect, relative to 2019.**

**The top 20 percent of Census Tracts with the most essential workers are outlined in black. At least 532 essential workers reside in each of these tracts. Combined, these tracts are home to 121,060 (41.5%) of the city's 293,575 essential workers.** "Essential workers" are defined as: healthcare practitioners and technical occupations; healthcare support occupations; food preparation and serving related occupations; building and grounds cleaning and maintenance occupations; and personal care and service occupations.

### Origin-Destination {.tabset .tabset-pills}

The spatial distribution of rideshare origins and destinations are very similar. Trips are highly concentrated around the downtown core - note the logarithmic scale of each map. While the number of trips in 2020 cratered compared to the same period in 2019, most trips started or ended around the central business district. In the Post-COVID periods of both 2019 and 2020, the top three community areas (The Loop, Near North Side, and Near West Side) accounted for over 40 percent of all rideshare trips within the city.

The widespread decline in the volume of trips, however, does mask some relative changes in neighborhood rideshare usage. The second map shows how the percentile rank of each Census Tract changed during the Post-COVID period in 2020 versus 2019. For example, one tract in the West Side neighborhood of Austin was middle of the pack for trip pickups in 2019 (56th percentile). Pickups were almost cut in half in 2020, but the tract qualified as one of the more active pickup spots in the city (75th percentile). Neighborhoods on the South and West Sides saw a relative increase in their percentile rank, indicating an inelastic demand for rideshare. While few rideshare trips were made in these areas compared to other parts of the city, rideshare trip behavior was less affected during the pandemic.

These South and West Side neighborhoods include some areas with a high concentration of essential workers, but the parts of the Southwest and Northwest Sides that saw the largest relative dropoff in their trip ranking are also home to many essential workers. 

#### Origin

```{r origin tracts tmap, fig.asp=1.5}

# Create essential worker definition
# National averages:
#     -Essential: 23.8%
#     -Healthcare: 9.5%
# City of Chicago:
#     -Essential: 334,381 (24.5%)
#     -Essential Non Police/Fire: 293,575 (21.5%)
#     -Healthcare: 118,787 (8.7%)
# Tract median (50th):
# Share of city employment in occupation
#     -Essential: 353 (0.11%)
#     -Essential Non Police/Fire: 314 (0.11%)
#     -Healthcare: 115 (0.10%)
# Tract top 20 percent, first quintile (80th):
# Share of city employment in occupation
#     -Essential: 610 (0.18%), combined 136,540 (40.8%)
#     -Essential Non Police/Fire: 532 (0.18%), combined 121,060 (41.5%)
#     -Healthcare: 214 (0.18%), combined 55,398 (46.6%)

# Quantile map
tmap_origin_facet

```

```{r origin tracts rank change tmap, fig.width=4}

# Percent rank change map
tmap_origin_rank_change +
  tm_view(set.view = 10.5,
          leaflet.options = list(zoomSnap = 0.5,
                                 zoomDelta = 0.5))

```

#### Destination

```{r destination tracts tmap, fig.asp=1.5}

# Quantile map
tmap_destination_facet

```

```{r destination tracts rank change tmap}

# Percent Rank Change
tmap_destination_rank_change

```

### Origin-Destination (OD) Flows

#### Flow Types

While origin and destination tracts provide a general sense of the geographic pattern of rideshare trips, they do not capture travel flows between areas. Before analyzing trip flows, it is important to distinguish between those trips that begin and end within the same tract (intra-flows) and trips that begin and end in different tracts (inter-flows). As seen in the table below, very few rideshare trips started and ended in the same tract. About 1.5 percent of all trips were intra-flows. Zooming out to the Community Area level, a far higher percentage of trips started and ended in the same neighborhood - roughly 20 percent. These figures were relatively consistent across years and time periods. 

```{r od intra vs inter kable}

# OD Pairs Intra vs Inter
# Pre/Post-COVID table
# Tracts
od_pairs_tracts_intra_inter_table %>%
  kbl(col.names = c("Trip Type", "2019", "2020", 
                    "2019", "2020", "Total")) %>% 
  add_header_above(c(" ", "Pre-COVID" = 2, "Post-COVID" = 2, " ")) %>%
  row_spec(3, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", 
                                      "condensed", "responsive"))

# Community Areas
od_pairs_ccas_intra_inter_table %>%
  kbl(col.names = c("Trip Type", "2019", "2020", 
                    "2019", "2020", "Total")) %>% 
  add_header_above(c(" ", "Pre-COVID" = 2, "Post-COVID" = 2, " ")) %>%
  row_spec(3, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", 
                                      "condensed", "responsive"))

```

#### Airport Trips

Across the entirety of 2019 and 2020, in both Pre-COVID and Post-COVID periods, the most popular trips were from O'Hare and Midway airports to the Loop, and vice versa. Trips to or from Chicago airports alone accounted for between 6% and 9% of all rideshare trips in each analysis period. These trips are excluded from the the OD flows maps below in order to focus on neighborhood-to-neighborhood travel patterns. 

```{r od tracts airport kable}

# OD Pairs Airport Trips
# Pre/Post-COVID table
od_pairs_airports_table %>% 
  kbl(col.names = c("Trip Type", "2019", "2020", 
                    "2019", "2020", "Total")) %>% 
  add_header_above(c(" ", "Pre-COVID" = 2, "Post-COVID" = 2, " ")) %>%
  row_spec(3, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", 
                                      "condensed", "responsive"))

```

#### OD Flows

The maps below link the beginning and end of trips (excluding airport trips), where the color and thickness of desire lines vary based on the number of trips made between Community Areas.

**The OD pairs indicate that the most common trips were short rides to and from the Loop and nearby areas like the West Loop and River North.** There is also a clear pattern of trips up and down the lakefront, particularly on the North Side. These trends are in line Streetsblog Chicago's [analysis of 2018 data](https://chi.streetsblog.org/2019/04/18/the-most-common-chicago-ride-hailing-trip-is-a-1-mile-hop-from-river-north-to-loop/). After the Stay-at-Home Order went into effect, trips linking the South lakefront to the Loop and the North Side decreased dramatically.

Neighborhoods with a high concentration of essential workers (532+, top 20 percent of Census Tracts) are outlined in black. 

```{r od flows ccas tmap, fig.asp=1.75}

# OD flows map (all community areas)
tmap_od_flows_ccas

```

## Drivers

The number of active drivers per month, defined as those making at least one trip during the reported month, declined significantly after March 2020, mirroring the decline in rideshare trips. This trend was also paired by a large increase in the number of inactive drivers, defined as those who completed zero trips despite being eligible to drive in Chicago that month. After the Stay-at-Home Order went into effect, a higher percentage of drivers were inactive (64%) compared to the same Post-COVID period in 2019 (34%).

The composition of drivers making trips did shift slightly in 2020 compared to 2019. A higher percentage of trips (+7%) were made by drivers with over a year's experience, both before and after the Stay-at-Home Order went into effect. Given that this trend was present during the early months of 2020 as well, it is unlikely that this was driven by the pandemic, but the shift is notable.

Almost all trips were made by drivers residing in Chicago (57%) or somewhere else in the State of Illinois (37%). Very few trips were made by drivers from other states. This pattern is steady over time, in both Pre- and Post-COVID periods.

### Active Drivers (1+ Trips)

```{r monthly active drivers interactive plot}

# Interactive plot
ggplotly(gg_monthly_drivers_active, tooltip = "text") %>% 
  layout(title = list(text = "Active Drivers per Month declined starting in March 2020, bottoming out at\n17,000 in April. The number of active drivers remained 50 percent\nbelow 2019 levels.",
                      y = 0.97))

```

### Driver Status

```{r drivers status interactive plot}

# Interactive plot
ggplotly(gg_drivers_status, tooltip = "text") %>% 
  layout(title = list(text = "After Stay-at-Home Order went into effect, a larger share of drivers were\ninactive (64%) compared to the same Post-COVID period in 2019 (34%).")) %>% 
  layout_ggplotly

```

### Trips by Driver's Tenure

```{r trips by drivers tenure interactive plot}

# Interactive plot
ggplotly(gg_drivers_tenure_trips, tooltip = "text") %>% 
  layout(title = list(text = "A higher percentage of trips (+7%) were made by drivers with over a\nyear's experience in 2020 compared to 2019, both before and after\nthe Stay-at-Home Order went into effect.",
                      y = 0.97)) %>% 
  layout_ggplotly

```

### Trips by Driver's License

```{r trips by drivers license interactive plot}

# Interactive plot
ggplotly(gg_drivers_license_trips, tooltip = "text") %>% 
  layout(title = list(text = "Most trips were made by drivers residing in Chicago (57%) or somewhere\nelse in the State of Illinois (37%). Very few trips were made by drivers\nfrom other states.",
                      y = 0.97)) %>% 
  layout_ggplotly

```