---
title: "Rideshare in Chicago During COVID-19"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    toc_depth: 4
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}

# Load libraries
library(tidyverse);library(lubridate);library(janitor)
library(sf);library(stplanr);library(leaflet)
library(leaflet.extras);library(tmap);library(gghighlight)
library(plotly);library(knitr);library(here)
library(kableExtra);library(viridis);library(scales)
library(hrbrthemes);library(ggthemes);library(RPostgres)
library(RPostgreSQL);library(DBI);library(config)

# Set working directory and options
setwd(here::here())
options(scipen = 999, stringsAsFactors = FALSE, dplyr.summarise.inform = FALSE)
knitr::opts_chunk$set(fig.width = 12, fig.asp = 0.618,
                      fig.align = "center", out.width = "100%")
tmap_mode("view")
tmap_options(basemaps = c(OSM = "OpenStreetMap", 
                          Light = "CartoDB.Positron",
                          Dark = "CartoDB.DarkMatter"))

```

## Rideshare Data

This analysis relies on Transportation Network Provider (TNP) data for trips, drivers, and vehicles. These datasets are published on the City of Chicago's Data Portal and updated quarterly. Please see the [GitHub repository](https://github.com/sean-connelly/Chicago_TNP) for source links and download instructions.

```{r import data}

# Connect to database
con <- RPostgres::dbConnect(RPostgres::Postgres(), 
                            host = config::get("host"),
                            port = config::get("port"),
                            dbname = config::get("dbname"),
                            user = config::get("user"), 
                            password = config::get("password"))

# Load in other data
# trips <- dbGetQuery(con, "SELECT * FROM public.trips LIMIT 100;")
# trips <- dbReadTable(con, "public.trips")

# analysis_trips <- dbGetQuery(con, "SELECT * FROM analysis.base_ytd LIMIT 100;")
# analysis_trips <- dbReadTable(con, "analysis.base_ytd")

# Drivers
# drivers <- dbGetQuery(con, "SELECT * FROM public.drivers LIMIT 100;")
# drivers <- dbReadTable(con, "public.drivers")

# Vehicles
# vehicles <- dbGetQuery(con, "SELECT * FROM public.vehicles LIMIT 100;")
# vehicles <- dbReadTable(con, "public.vehicles")

```

## Trips Over Time

### Daily Trips

```{r daily trips import}

# Import data
daily_trips <- dbGetQuery(con, "SELECT * FROM analysis.daily_trips")

# Clean and tidy
daily_trips <- daily_trips %>%
  mutate("year" = year(date),
         "day" = yday(date),
         "date_x" = date %>% 
           as.character(.) %>%
           str_replace("^\\d{4}","2000") %>% 
           as_date(.),
         "weekday" = wday(date, label = TRUE),
         "trips" = as.numeric(trips)) 

```



```{r daily trips plot}

# Plot information
daily_trips_plot_title <- daily_trips %>%
  group_by(year) %>%
  summarize(trips = sum(trips)) %>%
  pivot_wider(everything(), names_from = "year", values_from = "trips") %>% 
  mutate("change" = `2020`-`2019`,
         "change_pct" = (change/`2019`),
         "title_text" = paste0(as.character(comma(-change)), 
                               " (", percent(change_pct), ")"))

# Daily trips YoY plot
gg_daily_trips <- daily_trips %>% 
  group_by(year) %>% 
  arrange(date_x) %>%
  mutate(trips_agg = cumsum(trips)) %>% 
  ggplot(aes(x = date_x, y = trips_agg, 
             color = as.character(year),
             text = paste0("Date: ", paste0(format(date_x, "%m/%d"),
                                            "/", as.character(year)),
                           "<br>Agg. Trips: ", scales::comma(trips_agg)))) +
  geom_line(group = 1, size = 1.5) +
  geom_vline(xintercept = as.numeric(as.Date("2000-03-21")), 
             linetype = 4, color = "black", size = 1) +
  scale_x_date(date_breaks = "1 month", 
               date_labels = "%b") +
  scale_y_continuous(labels = comma,
                     limits = c(0, NA)) +
  labs(title = paste0("Compared to 2019, rideshare trips decreased by ",
                      daily_trips_plot_title$title_text, " in 2020"),
       x = "Day of the Year", y = "Aggregate Trips",
       color = "Year") +
  scale_color_ipsum() +
  theme_ipsum()

# Interactive plot
ggplotly(gg_daily_trips, tooltip = "text") %>% 
  layout(title = list(text =  paste0("Compared to 2019, rideshare trips decreased by ", daily_trips_plot_title$title_text, " in 2020")))

```

**This analysis uses the day that Governor Pritzker's Stay-at-Home Order went into effect (March 21st), as a way to delineate between two time frames:**

- **"Pre-COVID" (1/1-3/21) and **
- **"Post-COVID" (3/21-12/31)**

**As the data below suggest, the use of a single date to mark the onset of the pandemic is an oversimplification, but this allows for a clean comparison of rideshare trends year-over-year.** 

Prior to the Stay-at-Home Order going into effect, trips were already down by about 7 percent year-over-year. However, the bulk of the decline in rideshare occurred after the during the following 9 months. From late March 2020 onwards, only 27 million rideshare trips were taken in Chicago, about 60 million fewer trips compared to the same period in 2019.

```{r daily trips table}

# Pre/Post COVID table
daily_trips %>%
  mutate("covid_group" = ifelse(date_x >= "2000-03-21", 
                                "Post-COVID (3/21-12/31)", 
                                "Pre-COVID (1/1-3/21)")) %>% 
  mutate("covid_group" = factor(covid_group, 
                                levels = c("Pre-COVID (1/1-3/21)", 
                                           "Post-COVID (3/21-12/31)"))) %>% 
  group_by(covid_group, year) %>%
  summarize(trips = sum(trips)) %>%
  ungroup() %>% 
  pivot_wider(names_from = "year", 
              values_from = "trips") %>% 
  adorn_totals("row") %>% 
  mutate("change" = `2020`-`2019`,
         "change_pct" = (change/`2019`)) %>% 
  mutate(across(c(`2019`, `2020`, change), comma),
         "change_pct" = percent(change_pct)) %>% 
  select("Period" = covid_group, `2019`, `2020`, 
         "Change" = change, "Change (%)" = change_pct) %>% 
  kbl() %>% 
  row_spec(3, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", 
                                      "condensed", "responsive"))

```

### Monthly Trips

```{r monthly trips import}

# Import data
monthly_trips <- dbGetQuery(con, "SELECT * FROM analysis.monthly_trips")

# Clean and tidy
monthly_trips <- monthly_trips %>% 
  mutate("year" = year(month),
         "date_x" = floor_date(month, unit = "month") %>% 
           as.character(.) %>%
           str_replace("^\\d{4}","2000") %>% 
           as_date(.),
         "month" = month(month, label = TRUE, abbr = TRUE),
         "trips" = as.numeric(trips)) 

```

Taking a step back to examine trips per month, rideshare trips reached a low of 1.5 million in April. While trips rebounded in the following months, averaging around 3.6 million from July through October, they fell again in November and December as the country was entering the third wave of the pandemic.

```{r monthly trips}

# Monthly trips YoY plot
gg_monthly_trips <- monthly_trips %>% 
  ggplot(aes(x = date_x, y = trips, fill = as.character(year),
             text = paste0("Date: ", paste0(format(date_x, "%M"),
                                            "/", as.character(year)),
                           "<br>Trips: ", scales::comma(trips)))) +
  geom_bar(stat = "identity", color = "black",
           position = position_dodge(width = 25), width = 20) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  scale_y_continuous(labels = comma) +
  labs(title = "Trips per Month declined significantly in March, bottoming out at 1.5M in April.\nHowever, rideshare trips rebounded to 3.6M in the fall.",
       x = "Month", y = "Trips",
       fill = "Year") +
  scale_fill_ipsum() +
  theme_ipsum()

# Interactive plot
ggplotly(gg_monthly_trips, tooltip = "text") %>% 
  layout(title = list(text = "Trips per Month declined significantly in March, bottoming out at 1.5M in April.\nHowever, rideshare trips rebounded to 3.6M in the fall."))

```

### Trip Changes Surrounding 3/31/20 Stay-at-Home Order

It is clear that behavior changed in March even before the Stay-at-Home Order, as trips per day decreased significantly after Saturday, March 14th, 2020, a full week before the Order went into effect. Note that the earlier fluctuations in January and February correspond to weekday versus weekend trips.

```{r trips per day plot}

# Trips per day YoY plot
gg_trips_per_day <- daily_trips %>%
  filter(month(date, label = TRUE) %in% c("Feb", "Mar")) %>% 
  group_by(year) %>% 
  arrange(date_x) %>%
  ggplot(aes(x = date_x, y = trips, color = as.character(year),
             text = paste0("Date: ", paste0(format(date_x, "%m/%d"),
                                            "/", as.character(year)),
                           "<br>Trips: ", scales::comma(trips)))) +
  geom_line(group = 1, size = 1.5) +
  geom_vline(xintercept = as.numeric(as.Date("2000-03-21")), 
             linetype = 4, color = "black", size = 1) +
  scale_x_date(date_breaks = "1 week", 
               date_labels = "%b-%d") +
  scale_y_continuous(labels = comma,
                     limits = c(0, NA)) +
  labs(title = "Trips per Day fell before the Stay-at-Home Order went into effect,\n peaking at 365,000 trips on Saturday, March 14th, 2020",
       x = "Day of the Year", y = "Trips per Day",
       color = "Year") +
  scale_color_ipsum() +
  theme_ipsum()

# Interactive plot
ggplotly(gg_trips_per_day, tooltip = "text") %>% 
  layout(title = list(text = "Trips per Day fell before the Stay-at-Home Order went into effect,\npeaking at 365,000 trips on Saturday, March 14th, 2020"))

```

### Trips by Day of Week and Hour of Day

```{r day of week trips plot}

# Day of week YoY plot
# Pre/Post COVID table
gg_day_of_week_trips <- daily_trips %>%
  mutate("covid_group" = ifelse(date_x >= "2000-03-21", 
                                "Post-COVID", "Pre-COVID")) %>% 
  mutate("covid_group" = factor(covid_group, 
                                levels = c("Pre-COVID", "Post-COVID"))) %>% 
  group_by(covid_group, year, weekday) %>%
  summarize(trips = sum(trips)) %>%
  ungroup() %>%
  ggplot(aes(x = weekday, y = trips, fill = as.character(year))) +
  geom_bar(stat = "identity", color = "black") +
  # scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  scale_y_continuous(labels = comma) +
  labs(title = "Trips by Day of Week, Pre/Post-COVID",
       x = "Month", y = "Trips",
       fill = "Year") +
  facet_grid(covid_group ~ year) +
  scale_fill_ipsum() +
  theme_ipsum()

# Interactive plot
ggplotly(gg_day_of_week_trips) %>% 
  layout(title = list(text = "Trips by Day of Week, Pre/Post-COVID"))

```



```{r hour of day trips import}

# Import data
hourly_trips <- dbGetQuery(con, "SELECT * FROM analysis.hourly_trips")

# Clean and tidy
test <- hourly_trips %>%
  mutate("year" = year(date),
         "day" = yday(date),
         "date_x" = date %>% 
           as.character(.) %>%
           str_replace("^\\d{4}","2000") %>% 
           as_date(.),
         "weekday" = wday(date, label = TRUE),
         "trips" = as.numeric(trips)) 

```



```{r hour of day trips plot}

# Day of week YoY plot
# Pre/Post COVID table
gg_hour_of_day_trips <- daily_trips %>%
  mutate("covid_group" = ifelse(date_x >= "2000-03-21", 
                                "Post-COVID", "Pre-COVID")) %>% 
  mutate("covid_group" = factor(covid_group, 
                                levels = c("Pre-COVID", "Post-COVID"))) %>% 
  group_by(covid_group, year, weekday) %>%
  summarize(trips = sum(trips)) %>%
  ungroup() %>%
  ggplot(aes(x = weekday, y = trips, fill = as.character(year))) +
  geom_bar(stat = "identity", color = "black") +
  # scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  scale_y_continuous(labels = comma) +
  labs(title = "Trips by Hour of Day, Pre/Post-COVID",
       x = "Month", y = "Trips",
       fill = "Year") +
  facet_grid(covid_group ~ year) +
  scale_fill_ipsum() +
  theme_ipsum()

# Interactive plot
ggplotly(gg_hour_of_day_trips) %>% 
  layout(title = list(text = "Trips by Hour of Day, Pre/Post-COVID"))

```

## Trip Characteristics

```{r trip characteristics}

# Import data
trip_chars_avg <- dbGetQuery(con, "SELECT * FROM analysis.tnp_trip_chars") %>% 
  mutate("covid_group" = factor(covid_group, 
                                levels = c("Pre-COVID", "Post-COVID")),
         "trips" = as.numeric(trips)) %>% 
  arrange(covid_group, year)

```

### Distance

Prior to the Stay-at-Home Order going into effect, trips were 6.4 miles long on average, a 9 percent increase year-over-year. From late March 2020 onwards, however, the average trip distance increased to only slightly to 6.5 miles, a 1.7% increase compared to the same period in 2019.

```{r distance}

# Average distance
# Pre/Post COVID table
trip_chars_avg %>%
  select(covid_group, year, trip_miles) %>% 
  pivot_wider(names_from = "year", 
              values_from = "trip_miles") %>% 
  mutate("change" = `2020`-`2019`,
         "change_pct" = (change/`2019`)) %>% 
  mutate(across(c(`2019`, `2020`, change), comma),
         "change_pct" = percent(change_pct)) %>% 
  select("Period" = covid_group, `2019`, `2020`, 
         "Change" = change, "Change (%)" = change_pct) %>% 
  kbl(caption = "Average Trip Distance (Mi.)") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", 
                                      "condensed", "responsive"))

```

### Time

While average trip distance increased slightly, the average trip duration fell significantly in 2020. After March 21, 2020, the average trip time was 15.9 minutes, a 15% decrease compared to the same period in 2019. This finding aligns with Manzo and Bruno's [analysis based on a sample of trips from September 2019 and September 2020](https://illinoisepi.files.wordpress.com/2021/01/ilepi-pmcr-on-demand-workers-sub-minimum-wages-final.pdf) which found a 16% decline in average trip duration due to a decrease in congestion.

```{r time}

# Average time - minutes
# Pre/Post COVID table
trip_chars_avg %>%
  mutate("trip_minutes" = trip_seconds / 60) %>% 
  select(covid_group, year, trip_minutes) %>% 
  pivot_wider(names_from = "year", 
              values_from = "trip_minutes") %>% 
  mutate("change" = `2020`-`2019`,
         "change_pct" = (change/`2019`)) %>% 
  mutate(across(c(`2019`, `2020`, change), comma),
         "change_pct" = percent(change_pct)) %>% 
  select("Period" = covid_group, `2019`, `2020`, 
         "Change" = change, "Change (%)" = change_pct) %>% 
  kbl(caption = "Average Trip Duration (Min.)") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", 
                                      "condensed", "responsive"))

```

### Fares

The average trip fare in 2020 was about $0.80 higher in the first three months of the year compared to the same period in 2019, a 7.4% increase. However, in the period after the Stay-at-Home Order went into effect, the average trip fare in 2020 was about the same as in 2019.

```{r fares, warning=FALSE}

# Import data
fares <- dbGetQuery(con, "SELECT * FROM analysis.tnp_fares") %>% 
  mutate("covid_group" = factor(covid_group, 
                                levels = c("Pre-COVID", "Post-COVID")),
         "trips" = as.numeric(trips))

# Distribution of fares
fares_pct <- fares %>%
  group_by(covid_group, year) %>% 
  arrange(year, covid_group, fare) %>% 
  mutate("trips_pct" = trips / sum(trips),
         "trips_pct_agg" = cumsum(trips_pct)) %>% 
  ungroup()

# 95% of trips in all periods <$30 dollars
fares_tail <- fares_pct %>% 
  group_by(covid_group, year) %>% 
  filter(trips_pct_agg < 0.95) %>% 
  arrange(desc(fare)) %>% 
  slice(1) %>% 
  ungroup()

# Average fares
# Pre/Post COVID table
trip_chars_avg %>%
  select(covid_group, year, fare) %>% 
  pivot_wider(names_from = "year", 
              values_from = "fare") %>% 
  mutate("change" = `2020`-`2019`,
         "change_pct" = (change/`2019`)) %>% 
  mutate(across(c(`2019`, `2020`, change), dollar),
         "change_pct" = percent(change_pct)) %>% 
  select("Period" = covid_group, `2019`, `2020`, 
         "Change" = change, "Change (%)" = change_pct) %>% 
  kbl(caption = "Average Trip Fare ($)") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", 
                                      "condensed", "responsive"))
  
# Histogram
gg_fares <- fares %>%
  ggplot(aes(x = fare, y = trips, fill = as.character(year),
             text = paste0("Fare: ",scales::dollar(fare),
                           "<br>Trips: ", scales::comma(trips)))) +
  geom_bar(stat = "identity", color = "black") +
  geom_vline(data = trip_chars_avg, 
             mapping = aes(xintercept = fare),
             linetype = "dashed", size = 1) +
  scale_x_continuous(labels = dollar_format(prefix = "$"),
                     limits = c(0, max(fares_tail$fare))) +
  scale_y_continuous(labels = comma) +
  labs(title = "Fare Distribution to the Nearest $2.50, Pre/Post-COVID",
       subtitle = "95% of trip fares were less than $30 in each period",
       x = "Fare", y = "Trips",
       fill = "Year") +
  facet_grid(covid_group ~ year) +
  scale_fill_ipsum() +
  theme_ipsum()

# Interactive plot
ggplotly(gg_fares, tooltip = "text") %>% 
  layout(title = list(text = paste0("Fare Distribution to the Nearest $2.50, Pre/Post-COVID", 
                                    "<br>",
                                    "<sup>",
                                    "95% of trip fares were less than $30 in each period","</sup>")))

```

## Spatial Distribution of Trips

### Origin

```{r origin tracts data import}

# Import data
origin_tracts_comp <- dbGetQuery(con, "SELECT * FROM analysis.pickup_tracts_comp")
sf_tracts <- st_read(con, layer = "spatial_tracts")
acs_data_profiles <- dbGetQuery(con, "SELECT * FROM public.acs_data_profiles")

# Join trip data to spatial, reconfigure
origin_tracts_comp <- left_join(sf_tracts %>% select(GEOID),
                                origin_tracts_comp,
                                by = c("GEOID" = "pickup_census_tract")) %>% 
  st_as_sf() %>% 
  mutate("trips" = as.numeric(trips),
         "covid_group" = factor(covid_group, 
                                levels = c("Pre-COVID", "Post-COVID"))) %>% 
  arrange(covid_group, GEOID)

# Create essential worker definition
# National averages:
#     -Essential: 23.8%
#     -Healthcare: 9.5%
# Tract averages:
#     -Essential: 418 (26.5%)
#     -Healthcare: 148 (9.1%)
# Tract median (50th):
#     -Essential: 353 (26.1%)
#     -Healthcare: 115 (8.1%)
# Tract first quintile (80th):
#     -Essential: 610 (33.9%)
#     -Healthcare: 214 (12.6%)
acs_data_profiles <- acs_data_profiles %>% 
  filter(table_name == "S2401") %>% 
  select(GEOID, variable, estimate) %>% 
  pivot_wider(names_from = "variable",
              values_from = "estimate") %>% 
  mutate("essential" = select(., S2401_C01_015:S2401_C01_025) %>% 
           rowSums(na.rm = TRUE),
         "essential_pct" = essential / S2401_C01_001,
         "health_all" = select(., S2401_C01_015:S2401_C01_019) %>% 
           rowSums(na.rm = TRUE),
         "health_all_pct" = health_all / S2401_C01_001) %>% 
  rename("total" = S2401_C01_001,
         "health_pract" = S2401_C01_015,
         "health_support" = S2401_C01_019,
         "police_fire" = S2401_C01_020,
         "food_service" = S2401_C01_023,
         "custodial" = S2401_C01_024,
         "personal_service" = S2401_C01_025)

# Essential workers by tract
essential_workers_sf <- left_join(sf_tracts %>% select(GEOID),
                                acs_data_profiles,
                                by = "GEOID")

```

In the Pre-COVID period (1/1-3/21) the spatial distribution of rideshare pickups and dropoffs in 2019 and 2020 were very similar. After the Stay-at-Home Order went into effect, however, trips cratered, particularly those originating on the South, West, and Northwest Sides. Even the core area of high trip activity, roughly comprising the Loop and the North Side, shrunk in size.

**Census tracts where more than 33% of workers are essential are outlined in white, comprising the top 20 percent of tracts in the city. "Essential workers" are defined as: healthcare practitioners and technical occupations; healthcare support occupations; protective service occupations; food preparation and serving related occupations; building and grounds cleaning and maintenance occupations; and personal care and service occupations.** Neighborhoods with a high concentration of essential workers, especially those on the South and West Sides, saw a large decrease in trip dropoffs and pickups. 

```{r origin tracts map, fig.asp=1.75}

# Quantile map
tm_shape(origin_tracts_comp,
         name = "Tracts") +
  tm_fill(col = "trips", title = "Trips", alpha = 0.5,
          style = "quantile", palette = "viridis") +
  tm_borders() +
  tm_facets(by = c("covid_group", "year"),
            free.coords = FALSE,
            drop.units = TRUE) +
  tm_shape(essential_workers_sf %>% 
             filter(essential_pct >= quantile(essential_pct,
                                              0.80,
                                              na.rm = TRUE)),
           name = "Essential Workers") +
  tm_borders(col = "white", lwd = 2) +
  tm_layout(main.title = "Rideshare Trips by Origin Census Tract", 
            main.title.position = c("center"),
            legend.outside = FALSE,
            legend.width = 1.5,
            scale = 2.75) +
  tm_view(view.legend.position = c("left", "bottom"))

```

### Destinations

```{r destination tracts data import}

# Import data
destination_tracts_comp <- dbGetQuery(con, "SELECT * FROM analysis.dropoff_tracts_comp")

# Join tabular to spatial, reconfigure
destination_tracts_comp <- left_join(sf_tracts %>% select(GEOID),
                                     destination_tracts_comp,
                                     by = c("GEOID" = "dropoff_census_tract")) %>% 
  st_as_sf() %>% 
  mutate("trips" = as.numeric(trips),
         "covid_group" = factor(covid_group, 
                                levels = c("Pre-COVID", 
                                           "Post-COVID"))) %>% 
  arrange(covid_group, GEOID)

```



```{r destination tracts map, fig.asp=1.75}

# Quantile map
tm_shape(destination_tracts_comp,
         name = "Tracts") +
  tm_fill(col = "trips", title = "Trips", alpha = 0.5,
          style = "quantile", palette = "viridis") +
  tm_borders() +
  tm_facets(by = c("covid_group", "year"),
            free.coords = FALSE,
            drop.units = TRUE) +
  tm_shape(essential_workers_sf %>% 
             filter(essential_pct >= quantile(essential_pct,
                                              0.80,
                                              na.rm = TRUE)),
           name = "Essential Workers") +
  tm_borders(col = "white", lwd = 2) +
  tm_layout(main.title = "Rideshare Trips by Destination Census Tract", 
            main.title.position = c("center"),
            legend.outside = FALSE,
            legend.width = 1.5,
            scale = 2.75) +
  tm_view(view.legend.position = c("left", "bottom"))

```

### Origin-Destination (OD) Flows

```{r od tracts data import, warning=FALSE, message=FALSE}

# Import data
od_pairs_raw <- dbGetQuery(con, "SELECT * FROM analysis.trip_patterns_tracts_comp") %>% 
  mutate("covid_group" = factor(covid_group, 
                                levels = c("Pre-COVID", "Post-COVID")),
         "trips" = as.numeric(trips))

# Intra- vs inter-tract pairs
od_pairs_intra_inter <- od_pairs_raw %>% 
  mutate("intra_inter" = case_when(pickup_census_tract == 
                                     dropoff_census_tract ~
                                     "Intra (Within Same Tract)",
                                   pickup_census_tract != 
                                     dropoff_census_tract ~
                                     "Inter (Different Tract)")) %>% 
  group_by(covid_group, year, intra_inter) %>% 
  summarize("trips" = sum(trips)) %>% 
  ungroup()

```

#### Flow Types

While origin and destination tracts provide a general sense of the geographic pattern of rideshare trips, they do not capture travel flows between areas. Before analyzing trip flows, it is important to distinguish between those trips that begin and end within the same tract (intra-flows) and trips that begin and end in different tracts (inter-flows). As seen in the table below, very few rideshare trips started and ended in the same tract. About 1.5% of all trips were intra-flows. This was relatively consistent across years and time periods. 

```{r od tracts intra vs inter}

# OD Pairs Intra vs Inter
# Pre/Post-COVID table
od_pairs_intra_inter %>% 
  pivot_wider(names_from = c("year", "covid_group"),
              values_from = "trips") %>% 
  adorn_totals(c("row", "col")) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting() %>%
  kbl(col.names = c("Trip Type", "2019", "2020", 
                    "2019", "2020", "Total")) %>% 
  add_header_above(c(" ", "Pre-COVID" = 2, "Post-COVID" = 2, " ")) %>%
  row_spec(3, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", 
                                      "condensed", "responsive"))

```



```{r od tracts airport data import}

# Calculate airport trips
od_pairs_airports <- od_pairs_raw %>% 
  filter(pickup_census_tract != dropoff_census_tract,
         pickup_census_tract %in% sf_tracts$GEOID,
         dropoff_census_tract %in% sf_tracts$GEOID) %>% 
  mutate("airports" = ifelse(
    pickup_census_tract %in% c("17031980000", "17031980100") | dropoff_census_tract %in% c("17031980000", "17031980100"), 
    "To or From O'Hare/Midway",
    "Other Trip")) %>% 
  group_by(covid_group, year, airports) %>% 
  summarize("trips" = sum(trips)) %>% 
  ungroup()

```

#### Airport Trips

Across the entirety of 2019 and 2020, in both Pre-COVID and Post-COVID periods, the most popular trips were from O'Hare and Midway airports to the Loop, and vice versa. Trips to or from Chicago airports alone accounted for between 6% and 9% of all rideshare trips in each analysis period. These trips are excluded from the the OD flows maps below in order to focus on neighborhood-to-neighborhood travel patterns. 

```{r od tracts airport}

# OD Pairs Airport Trips
# Pre/Post-COVID table
od_pairs_airports %>% 
  pivot_wider(names_from = c("year", "covid_group"),
              values_from = "trips") %>% 
  adorn_totals(c("row", "col")) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting() %>%
  kbl(col.names = c("Trip Type", "2019", "2020", 
                    "2019", "2020", "Total")) %>% 
  add_header_above(c(" ", "Pre-COVID" = 2, "Post-COVID" = 2, " ")) %>%
  row_spec(3, bold = TRUE) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", 
                                      "condensed", "responsive"))

```



```{r od desire lines}

# Remove airport trips, make sure in Chicago
# Convert into desire lines for mapping
od_pairs_comp <- od_pairs_raw %>% 
  filter(pickup_census_tract != dropoff_census_tract,
         pickup_census_tract %in% sf_tracts$GEOID,
         dropoff_census_tract %in% sf_tracts$GEOID,
         !pickup_census_tract %in% c("17031980000", "17031980100"),
         !dropoff_census_tract %in% c("17031980000", "17031980100")) %>%
  select(pickup_census_tract, dropoff_census_tract,
         everything()) %>% 
  od2line(flow = ., zones = sf_tracts)

# Create a linewidth variable to scale desire lines
od_pairs_comp <- od_pairs_comp %>% 
  mutate("lwd" = trips / mean(od_pairs_comp$trips)) %>% 
  relocate(lwd, .after = trips)

```

#### OD Flows, 1000+ Trips

The maps below link the beginning and end of trips (excluding airport trips), where the color and thickness of desire lines vary based on the number of trips made between tracts. 

The OD pairs indicate a high concentration of trips linking the Central Business District to nearby areas like the West Loop, as well as a clear pattern of trips up and down the lakefront. After the Stay-at-Home Order went into effect, trips linking the South lakefront to the Loop and the North Side decreased dramatically.

**Neighborhoods with a high concentration of essential workers (33%+) are outlined in black.** 

```{r origin-destination tracts map, eval=FALSE, fig.asp=1.75}

# Distribution of OD trip pairs
od_pairs_pct <- od_pairs_comp %>%
  st_drop_geometry() %>% 
  group_by(covid_group, year) %>% 
  arrange(year, covid_group, desc(trips)) %>% 
  mutate("trips_pct" = trips / sum(trips),
         "trips_pct_agg" = cumsum(trips_pct)) %>% 
  ungroup()

# 90% of OD trip pairs in all periods >10 trips
od_pairs_tail <- od_pairs_pct %>% 
  group_by(covid_group, year) %>% 
  filter(trips_pct_agg < 0.90) %>% 
  arrange(trips) %>% 
  slice(1) %>% 
  ungroup()

# OD flows map (all tracts)
tm_shape(sf_tracts) +
  tm_borders() +
  tm_shape(od_pairs_comp %>% 
             filter(trips >= 1000),
           name = "OD Flows") +
  tm_lines(col = "trips", title = "Trips", 
           style = "quantile", palette = "viridis",
           lwd = "lwd", title.lwd = "Trips",
           scale = 40, alpha = 0.5,
           legend.lwd.show = FALSE) +
  tm_facets(by = c("covid_group", "year"), 
            free.coords = FALSE,
            drop.units = TRUE) +
  tm_shape(essential_workers_sf %>% 
             filter(essential_pct >= quantile(essential_pct,
                                              0.80,
                                              na.rm = TRUE)),
           name = "Essential Workers") +
  tm_borders(col = "black", lwd = 2) +
  tm_layout(main.title = "Rideshare Trips by Origin-Destination (OD) Pairs\nOD Pairs with 1,000+ Trips", 
            main.title.position = c("center"),
            legend.outside = FALSE,
            legend.width = 1.5,
            scale = 2.75) +
  tm_view(view.legend.position = c("left", "bottom"))

```




## Drivers

```{r}

```

